Quantifying the home advantage with `hyper3` objects

### 2023-01-30 inst/home_advantage3.Rmd
rewrite

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
$$

@@

$$
\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}
$$

```{r label=defineABCD}
A <- 9
B <- 2
C <- 3
D <- 5
```

$$
\hat{p_1}=\frac{\sqrt{27}}{\sqrt{27} + \sqrt{10}}\simeq 0.62\qquad
\hat{p_2}=\frac{\sqrt{15}}{\sqrt{27} + \sqrt{10}}\simeq 0.38\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}\simeq 2.73$$

Keeping $A=x,B=x,C=x,D=xxxxxx$ and $\lambda=2.73$ [assumed for the moment
to be known, estimation techniques are discussed later], appropriate
package idiom might be to use bespoke function `home_away3()`:

```{r,label=usehomeaway3a}
library("hyper2",quietly=TRUE)
M <- matrix(c(NA, A+B*1i ,D+C*1i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
(H <- home_away3(M,lambda=2.73))
```

Above, obect `M` has real parts being home wins and imaginary parts
being away wins.

We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
package idiom:

```{r,label=findtheevaluate3,cache=TRUE}
H
maxp(H)
```

@@

```{r,label=estimatelambda,cache=TRUE}
f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
f(2)
f(2.1)
lam <- seq(from=1,to=10,len=17)
Supp <- sapply(lam,f)
```

@@

```{r,label=optimlam,cache=TRUE}
op <- optimize(f,interval=c(2,3),maximum=TRUE)
op
```

@@

```{r,ept,cache=TRUE}
equalp.test(home_away3(M,lambda=op$maximum))
```

>>>>

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\underbrace{
\overbrace{\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
\overbrace{\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
\underbrace{
\overbrace{\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
\overbrace{\left(\frac{        p_1}{p_1 + \lambda p_2}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
$$

@@

$$
\hat{p_1}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
\hat{p_2}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
\hat{\lambda}=\sqrt{\frac{AC}{BD}}
$$

Note that the evaluate is unchanged if $A,B,C,D$ are all multiplied by
a positive constant.

## Aside: `hyper2` analysis of the same scoreline

Now the likelihood function would be

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\underbrace{
\overbrace{\left(\frac{p_1+M}{p_1 + p_2+M}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
\overbrace{\left(\frac{p_2  }{p_1 + p_2+M}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
\underbrace{
\overbrace{\left(\frac{p_2+M}{p_1 + p_2+M}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
\overbrace{\left(\frac{p_1  }{p_1 + p_2+M}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
$$

The maximum would be

$$
\hat{p_1}=\frac{C}{C+D}\qquad
\hat{p_2}=\frac{A}{A+B}\qquad
M=\frac{AD-BC}{(A+B)(B+C)}
$$

although it is not clear to me what the _constrained_ optimal point
[that is, requiring $M\geq 0$] is.

## Numerical analysis

```{r label=defineABCD}
A <- 9
B <- 5
C <- 7
D <- 3

(p1hat <- sqrt(A*D)/(sqrt(A*D) + sqrt(B*C)))
(p2hat <- sqrt(B*C)/(sqrt(A*D) + sqrt(B*C)))

(lam_hat <- sqrt((A*C)/(B*D)))
```

$$
\hat{p_1}=\frac{\sqrt{21}}{\sqrt{21} + \sqrt{35}}\simeq 0.47\qquad
\hat{p_2}=\frac{\sqrt{63}}{\sqrt{21} + \sqrt{35}}\simeq 0.53\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}=2.05.
$$

Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
to be known, estimation techniques are discussed later], we can
represent this information in a standard format:

```{r,label=usehomeaway3a}
library("hyper2",quietly=TRUE)
M <- matrix(c(NA, A+B*1i ,C+D*1i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
```

Above, object `M` has real parts being home wins and imaginary parts
being away wins.  Appropriate package idiom to specify a likelihood
function might be to use bespoke function `home_away3()`:

```{r,label=ha3calc,cache=TRUE}
(H <- home_away3(M,lambda=2.05))
```

Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
to be known, estimation techniques are discussed later], we may
estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in package
idiom:

```{r,label=findtheevaluateh3,cache=TRUE}
H
maxp(H)
```

@@

```{r,label=estimatelambda3,cache=TRUE}
f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
f(2)
f(2.1)
lam <- seq(from=1,to=10,len=17)
Supp <- sapply(lam,f)
```

@@

```{r,label=optimlambda,cache=TRUE}
op <- optimize(f,interval=c(2,3),maximum=TRUE)
op
```

@@

```{r,ept3,cache=TRUE}
equalp.test(home_away3(M,lambda=op$maximum))
```
